---
title: "Understanding and Modifying ODE Structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding and Modifying ODE Structures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(ndm)
```

This vignette is maintainer-facing. It explains how ODE structures are
represented in `ndm`, how the package distinguishes built-in versus structured
families, what the TeX parser expects, and how to modify or extend structures
without changing the runtime contract by accident.

## Mental model

`ndm_model_spec` is the canonical package-side representation of an ODE
structure. You can obtain one in three ways:

1. Select a built-in preset with `ndm_model_spec()`.
2. Build a structured declaration with `ndm_model_spec_from_structure()`.
3. Import a custom TeX file with `ndm_model_spec_from_tex()`.

All three routes eventually produce the same kind of object. The legacy runtime
builder still consumes TeX, so structured declarations are converted to TeX
before the NeuralODE parser sees them.

```{r}
spec <- ndm_model_spec(preset = "seirs_dynamic_beta", model_type = "NeuralODE")
class(spec)
print(spec)
```

## Built-in taxonomy

Use `ndm_model_spec_presets()` to inspect the packaged presets and their high
level metadata.

```{r}
presets <- ndm_model_spec_presets()
presets[, c("preset", "compartments", "dynamic_beta", "dynamic_global", "multi_outcome")]
```

There are two broad families:

- Built-in SEIR/SEIRS presets are registered in `R/specs.R` and point at TeX
  source files under `tools/runtime_source/ndm_runtime/ModelStructureTex`.
- Structured TB families are registered in `R/specs_structured.R` and expand
  parameterized families into explicit equations before they are serialized to
  TeX.

The metadata columns are meant as quick selectors rather than a full semantic
description:

- `dynamic_beta`: local transmission terms like `beta_l` or related latent
  local dynamics are time-varying.
- `dynamic_global`: global rates such as `sigma`, `gamma`, `xi`, or `delta`
  evolve through a global neural block.
- `multi_outcome`: more than one `\\Observe` expression is emitted.
- `family_args`: extra arguments for structured families such as chain length
  in `tb_b` or `tb_j`, and TB observation semantics.
- `time_varying_terms` and `endogenous_terms`: metadata used to describe the
  dynamic terms on the spec object; inspect them directly on the returned spec.

```{r}
tb_spec <- ndm_model_spec(
  preset = "tb_b",
  model_type = "NeuralODE",
  family_args = list(n = 4L)
)

tb_spec$family
tb_spec$family_args
tb_spec$state_terms
```

Structured TB presets keep `i_l` as the cumulative endpoint for progression to
active TB. Their default observation is the incidence flow into `i_l`; use
`family_args = list(observation_target = "cumulative")` only when the data target
is cumulative progression risk rather than incident active TB.

### Dynamics variants

Every `tb_*` family can be expanded under one of three `dynamics_variant`
settings passed via `family_args`:

- `"progression"` — the raw one-way depletion equations from
  `reports/TBModels/TBDiseaseTypes.md` (constant per-capita force of infection
  `lambda * s_l`, no removed compartment).
- `"balanced_incidence"` (default) — the progression topology plus demographic
  renewal (`mu`), active-disease exit back to susceptibility (`gamma_i`), and
  latent mortality, so incidence stays sustained for fitting real panels.
- `"seir"` — the SEIR-style form of
  `reports/TBModels/TBDiseaseTypes_v2.md`. The constant force of infection is
  replaced by mass-action incidence `beta * i_l / N` (the parameter `lambda` is
  renamed to `beta`); an explicit removed compartment `r_l` is added with
  active-disease removal `gamma` (I to R); and demographic renewal `mu` keeps
  the epidemic endemic rather than a single burn-out wave. The observation
  remains the gross active-TB incidence inflow (recorded before removal), so for
  most structures it is unchanged, while the degenerate direct-progression
  structure D and the direct-progression fractions of E and G observe the
  mass-action inflow itself.

```{r seir-variant, eval = FALSE}
seir_spec <- ndm_model_spec(
  preset = "tb_a",
  model_type = "NeuralODE",
  family_args = list(dynamics_variant = "seir")
)
seir_spec$state_terms                     # s_l, l_l, i_l, r_l
seir_spec$execution_spec$equations[["r_l"]]  # "gamma * i_l - mu * r_l"
seir_spec$execution_spec$observations     # "c * l_l" (gross incidence inflow)
```

## Structured declarations

`ndm_model_spec_from_structure()` accepts either:

- a short declaration with `family = "tb_*"` plus optional `family_args`, or
- a fully explicit declaration containing `states`, `parameters`, `equations`,
  and `observations`.

The explicit form is useful when you want a new structure without editing the
package registries first.

```{r}
toy_structure <- list(
  preset = "toy_slqi",
  description = "Toy four-state latent structure with a holding compartment.",
  states = c("s_l", "l_l", "q_l", "i_l"),
  parameters = list(
    lambda = list(
      transformation = "InvSoftPlus",
      prior_mean = 0.08,
      prior_sd = 0.25
    ),
    c = list(
      transformation = "InvSoftPlus",
      prior_mean = 0.05,
      prior_sd = 0.25
    ),
    h = list(
      transformation = "InvSoftPlus",
      prior_mean = 0.03,
      prior_sd = 0.25
    )
  ),
  equations = c(
    s_l = "- lambda * s_l",
    l_l = "lambda * s_l - (c + h) * l_l",
    q_l = "h * l_l",
    i_l = "c * l_l"
  ),
  observations = "i_l",
  init_state_terms = c("s_l", "l_l", "q_l", "i_l"),
  neural = list(local_inputs = c("s_l", "l_l", "q_l", "i_l"))
)

toy_spec <- ndm_model_spec_from_structure(
  toy_structure,
  model_type = "NeuralODE"
)

toy_spec$state_terms
toy_spec$parameter_terms
```

For direct structured declarations, each parameter is currently represented as
a named list with these fields:

- `transformation`: one of `Identity`, `InvSoftPlus`, or `InvSigmoid`
- `prior_mean`
- `prior_sd`

That is the current contract consumed by the TeX serializer. It is not wrapped
in a public helper yet, so maintainers should document any future changes to
that shape if the API is formalized later.

## TeX contract

The runtime parser still works from TeX snippets bounded by explicit markers:

```{r}
tex_lines <- strsplit(ndm_model_spec_to_tex(spec), "\n", fixed = TRUE)[[1L]]
tex_lines[grep(
  "%%%START BAYES ODE%%%|%%%END BAYES ODE%%%|Evolve|Observe|sim|leftarrow",
  tex_lines
)]
```

The important conventions are:

- `%%%START BAYES ODE%%%` and `%%%END BAYES ODE%%%` mark the parseable block.
- `\\item $x \\leftarrow ...$` defines constants.
- `\\item $... \\sim \\Dist{Normal}(... )$` defines priors.
- Init-state priors are distinguished by names like `s_{l0}` and are mapped
  back to `s_l`, `e_l`, and so on.
- `\\Evolve{state}` defines one ODE equation per evolved term.
- `\\Observe = ...` defines one or more observation expressions.
- `\\Neural1{...}` denotes local dynamics.
- `\\Neural2{...}` denotes global dynamics.

In the current runtime code, these conventions are parsed mostly through string
processing. When you change naming or formatting, assume the parser is strict
unless tests prove otherwise.

## Safe modification workflow

When you want to modify an existing structure without creating a new package
family, start from a packaged spec, round-trip it through TeX, and inspect the
resulting metadata before using it in training code.

```{r}
base_spec <- ndm_model_spec(
  preset = "seirs_dynamic_beta_dynamic_global",
  model_type = "NeuralODE"
)

tmp <- tempfile(fileext = ".tex")
ndm_model_spec_to_tex(base_spec, tmp)
roundtrip <- ndm_model_spec_from_tex(tmp, model_type = "NeuralODE")

identical(base_spec$state_terms, roundtrip$state_terms)
identical(base_spec$init_state_terms, roundtrip$init_state_terms)
identical(base_spec$parameter_terms, roundtrip$parameter_terms)
```

Use this route when you are experimenting with:

- reordered `\\Evolve{}` blocks
- alternative `\\Observe` expressions
- new dynamic terms in an otherwise existing model

If you are only varying arguments of an existing structured family, prefer
`family_args` rather than editing TeX directly.

## Extending the package

Choose the extension route that matches the change:

### Add a built-in TeX preset

Do this when the structure is best expressed as a maintained named preset.

- Add or update the TeX source under
  `tools/runtime_source/ndm_runtime/ModelStructureTex`.
- Register the preset in `.ndm_builtin_spec_registry()` in `R/specs.R`.
- Update any default `time_varying_terms` or `endogenous_terms` mappings in
  `ndm_model_spec()`.
- Refresh generated runtime artifacts before documenting or building the
  package. `documentPackage.R` already runs the generation helpers; if working
  manually, run `tools/generate_runtime_registry.R` and
  `tools/generate_runtime_sources.R`.

### Add a structured family

Do this when the family is parameterized or easier to maintain as equations in
R than as hand-written TeX.

- Add the family builder logic in `.ndm_tb_structure_spec()` or the equivalent
  structured-family constructor.
- Register the family in `.ndm_structured_spec_registry()`.
- Confirm the resulting spec emits the right TeX metadata and round-trips back
  through `ndm_model_spec_from_tex()`.

## Validation checklist

Before trusting a modified or new structure:

1. Create the spec and print it to confirm `preset`, `family`, and
   `source_format`.
2. Check `state_terms`, `init_state_terms`, and `parameter_terms`.
3. Round-trip through TeX with `ndm_model_spec_to_tex()` and
   `ndm_model_spec_from_tex()`.
4. Run the spec-focused tests in `tests/testthat/test-specs.R`.
5. If the structure is meant to execute end to end, run the relevant
   NeuralODE smoke coverage in `tests/testthat/test-sim-fit.R`.

Those tests currently provide the most concrete examples of supported
modification patterns, especially for custom init-state layouts and the TB
family coverage.
